\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb}
\usepackage{url}

\usepackage{pl-judgments}

\title{\texttt{pl-judgments} Package%
\footnote{\copyright{} \the\year{} Robert Harper.  All Rights Reserved.}}
\author{Robert Harper}
\date{\today}

\begin{document}

\maketitle{}

\section*{Overview}

Include the declaration \verb|\usepackage{pl-judgments}| to define macros for the judgment forms that are typically used when defining programming languages.\footnote{The former package name, \texttt{pfpl-judgments}, is denigrated, but may still be used.}  See the source code for this document for the names of the macros and their use.

\section*{Statics}
\label{sec:statics}

\subsection*{Basic Judgments}
\label{sec:statics-basic}

Languages with a phase separation distinguish types, which classify values, from expressions, which evaluate to values.  Languages with effects have a \emph{modal separation} between expressions and commands, the latter of which are classified by the types of their returned values.  Languages with \emph{explicit control} classify contexts aka continuations by the type of value that they accept.

In richer languages there are non-trivial notions of equality of types, and correspondingly equality of constructors, and in languages with dependent kinds, there is also equality of kinds.  In all cases classification respects equality of classifiers.  In some cases there is also a notion of subtyping, and singleton kinds induce forgetful subkinding relations.  Subsumption specifies that typing and kinding respect these relations.

The basic judgment forms corresponding to this architecture are summarized in Figure~\ref{fig:statics}.

\subsection*{Hypothetical/General Judgments}
\label{sec:statics-hypgen}

Hypothetical/general judgments have the form $\Gamma\entails J$, which is written \verb|\Gamma\entails J|, where $J$ is a basic judgment, and $\Gamma$ is a context of the form
\[
    j_1,\dots,j_n
\]
where $n\geq 0$, and the $j_i$'s are variable declarations whose form depends on the language being considered.  As such contexts are inductively generated from the empty context, written \verb|\empCtx|, by context extension, written \verb|\extCtx{\Gamma}{j}|, to obtain $\extCtx{\Gamma}{j}$.  The concatenation of two contexts is written \verb|\appCtx{\Gamma_1}{\Gamma_2}| to obtain $\appCtx{\Gamma_1}{\Gamma_2}$, and is defined in the evident way.

% In practice the type formation judgment, $\isTp{\tau}$, may be extended to the hypothetical form $\Delta\entails\isTp{\tau}$, where $\Delta$ is a context of the form
% \[
%     \extCtx{\isTp{t_1}\dots}{\isTp{t_n}}
% \]
% declaring that the $t_i$'s are type variables.  When kinds are involved, such a context has the form
% \[
%     \extCtx{\isOfKd*{u_1}{\kappa_1}\dots}{\isOfKd*{u_n}{\kappa_n}}.
% \]
% Similarly, the expression typing judgment may be extended to the hypothetical form $\Gamma\entails \isOf{e}{\tau}$, where $\Gamma$ is a context of the form
% \[
%     \extCtx{\isOf*{x_1}{\tau_1}\dots}{\isOf*{x_n}{\tau_n}}
% \]
% where the $x_i$'s are expression variables, and the $\tau_i$'s are types.  When both forms of variable are involved in an expression typing, the combined hypothetical form is written $\sepCtx{\Delta}{\Gamma}\entails J$, emphasizing that $\Delta$ is prior to $\Gamma$.

Separately from variables, which are unknowns of some syntactic category, there may also be \emph{symbols}, such as the names of assignables, to which a type is \emph{associated}.  These are declared using \emph{signatures}, $\Sigma$, of the form
\[
    \hasTp*{a_1}{\tau_1},\dots,\hasTp*{a_n}{\tau_n}
\]
where $n\geq 0$, and the $a_i$'s are distinct symbols, which stand only for themselves and which may be compared for equality and disequality.  As with contexts, signatures are generated by the empty signature, \verb|\empSig|, and signature extension, \verb|\extSig|, and may be combined using \verb|\appSig|.

% Thus, the statics of commands is then given by hypothetical judgments of the form $\Gamma\entails[\Sigma] \isOf{e}{\tau}$ and $\Gamma\entails[\Sigma] \hasTp{m}{\tau}$, where $m$ is a command.

\smallskip

A collection of formation and membership judgments is given in Figure~\ref{fig:statics}.  The binary classification judgments have ``starred'' forms for declaring variables for which the membership symbol is typeset slightly tighter (as a binary operation, rather than as a relation) for use in a context or signature.  Those that admit a starred form are marked as such in the usage column; the star should be omitted when used relationally.

\begin{figure}[p]
    \begin{displaymath}
        \begin{array}{l@{\qquad} l@{\qquad}l}
            \text{Judgment}   & \text{Meaning}  & \text{Invocation} \\
            \isTp{\tau}            & \tau\ \text{is a type} & \verb|\isTp{\tau}| \\
            \eqTp{\tau}{\tau'}     & \tau\ \text{and}\ \tau'\ \text{are equal types} & \verb|\eqTp{\tau}{\tau'}| \\
            \subTp{\tau}{\tau'}    & \tau\ \text{is a subtype of}\ \tau' & \verb|\subTp{\tau}{\tau'}| \\\\
            \isKd{\kappa}          & \kappa\ \text{is a kind} & \verb|\isKd{\kappa}| \\
            \eqKd{\kappa}{\kappa'} & \kappa\ \text{and}\ \kappa'\ \text{are equal kinds} & \verb|\eqKd{\kappa}{\kappa'}| \\
            \subKd{\kappa}{\kappa'} & \kappa\ \text{is a subkind of}\ \kappa' & \verb|\subKd{\kappa}{\kappa'}| \\\\
            \isOfKd{c}{\kappa}   & c\ \text{has kind}\ \kappa & \verb|\isOfKd*{c}{\kappa}|, \verb|\ofKd*{c}{\kappa}| \\
            \eqOfKd{c}{c'}{\kappa} & c\ \text{and}\ c'\ \text{are equal at kind}\ \kappa & \verb|\eqOfKd{c}{c'}{\kappa}| \\\\
            \isOfTp{e}{\tau}         & e\ \text{has type}\ \tau & \verb|\isOfTp*{e}{\tau}|, \\
            & & \verb|\isOf*{e}{\tau}|, \verb|\ofTp*{e}{\tau}| \\
            \retsTp{m}{\tau}       & m\ \text{returns type}\ \tau & \verb|\retsTp{m}{\tau}| \\
            \accsTp{k}{\tau}       & k\ \text{accepts type}\ \tau  & \verb|\accsTp{k}{\tau}| \\\\
            \isCtx{\Gamma}         & \Gamma\ \text{is a valid context} & \verb|\isCtx{\Gamma}| \\
            \isSig{\Sigma}         & \Sigma\ \text{is a valid signature} & \verb|\isSig{\Sigma}|
        \end{array}
    \end{displaymath}

    \caption{Judgments for Statics}
    \label{fig:statics}
\end{figure}

\section*{Dynamics}
\label{sec:dynamics}

Often the dynamics of languages is formulated using \emph{state transition systems}, which define a class of execution states, some of which are deemed initial, some of which are deemed final, and a relation between states that expresses ``one step'' of execution.  In principle a state can be anything at all, but in practice states are constructed from (closed, well-formed) pieces of syntax, perhaps involving symbols, and sometimes involving auxiliary data structures such as memories.  These considerations lead to the judgment forms given in Figure~\ref{fig:dynamics}.

There are, invariably, unlabelled forms of transition that may, or may not, be indexed by a signature defining the symbols that are available within the state.  In some cases labelled forms of transition are further indexed by an \emph{action} that specifies the effect of a step of execution.  Generally speaking, such transitions are not execution steps \textit{per se}, but are rather potential steps that may be taken in conjunction with another transition undertaking the \emph{complementary} action.

Occasionally it is helpful to introduce evaluation relations for expressions and execution relations for commands that consolidate multiple individual steps of execution to a final state.  These are also specified in Figure~\ref{fig:dynamics}.  In most situations these relations are secondary to the transition systems that define a dynamics, but in some situations they may be defined directly, without or without reference to a transition system.

\begin{figure}[p]
    \begin{displaymath}
        \begin{array}{l@{\qquad}l@{\qquad}l}
            \text{Judgment} & \text{Meaning} & \text{Invocation} \\
            \okSt[\Sigma]{s}       & s\ \text{is an ok state} & \verb|\okSt[\Sigma]{s}| \\
            \iniSt[\Sigma]{s}            & s\ \text{is initial}\ (\text{rel.} \Sigma) & \verb|\iniSt[\Sigma]{s}|, \verb|\isIni[\Sigma]{s}| \\
            \finSt[\Sigma]{s}            & s\ \text{is final}\ (\text{rel.} \Sigma) & \verb|\finSt[\Sigma]{s}|, \verb|\isFin[\Sigma]{s}| \\\\
            %
            s\stepsTo[\Sigma] s' & s\ \text{steps to}\ s'\ (\text{rel.}\ \Sigma) & \verb|s\stepsTo[\Sigma] s'| \\
            s\stepsTo[\Sigma](k) s' & s\ \text{steps to}\ s'\ \text{in}\ k\ \text{steps} & \verb|s\stepsTo[\Sigma](k) s'| \\
            s\stepsTo[\Sigma](\ast) s' & s\ \text{steps to}\ s'\ \text{in zero or more steps} & \verb|s\stepsTo[\Sigma](\ast)| \\\\
            %
            s\stepsTo[\Sigma]<\alpha> s' & s\ \text{steps to}\ s'\ \text{with action}\ \alpha\ (\text{rel.}\ \Sigma) & \verb|s\stepsTo[\Sigma]<\alpha> s'| \\\\
            %
            \isVal[\Sigma]{e}    & e\ \text{is a value}\ (\text{rel.}\ \Sigma) & \verb|\isVal[\Sigma]{e}| \\\\
            e\evalsTo[\Sigma] v          & e\ \text{evaluates to}\ v\ (\text{rel.}\ \Sigma) & \verb|e\evalsTo[\Sigma] v| \\
            e\evalsTo[\Sigma](c) v       & e\ \text{evaluates to}\ v\ \text{with cost}\ c & \verb|e\evalsTo[\Sigma](c) v| \\\\
            %
            e\raisesUp[\Sigma] v           & e\ \text{raises}\ v & \verb|e\raisesUp[\Sigma] v| \\
            e\raisesUp[\Sigma](c) v        & e\ \text{raises}\ v\ \text{with cost}\ c & \verb|e\raisesUp[\Sigma](c) v| \\\\
            %
            e\hasResult[\Sigma] v    & e\ \text{evaluates or raises}\ v & \verb|e\hasResult[\Sigma] v| \\
            e\hasResult[\Sigma](c) v    & e\ \text{evaluates or raises}\ v\ \text{with cost}\ c  & \verb|e\hasResult[\Sigma](c) v|
        \end{array}
    \end{displaymath}

    \caption{Judgments for Dynamics}
    \label{fig:dynamics}
\end{figure}

% \section*{Equality}
% \label{sec:equality}

% \subsection*{Formal Equality}
% \label{sec:equality-formal}

% \subsection*{Semantic Equality}
% \label{sec:equality-semantic}

\end{document}

